---
output: 
  pdf_document:
    keep_tex: yes
header-includes:
    - \input{report-template}
    - \input{latex-macros}
---

\printtitle


\section{Introduction}
some cool text.

You can use shorTeX for LaTeX macros in R:
https://github.com/trevorcampbell/shortex.

\newpage

lorem ipsum


```{r preamble, include=FALSE, message=FALSE, echo=FALSE, appendix=TRUE}
# R setup ------------------------------------#
library(knitr)
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(tidy = FALSE)
library(tidyverse)
library(ggplot2)
ggplot2::theme_set(theme_classic())
library(palmerpenguins)
library(kableExtra)
#--------------------------------------------#

```



# Tables with kable and kableExtra

Now let's do some nice tables.

```{r table1, appendix=TRUE}
# table 1 ------------------------#
table <- 
  penguins %>% 
  tidyr::drop_na() %>% 
  dplyr::group_by(species,sex) %>% 
  dplyr::summarise(mean_body_mass_g = mean(body_mass_g)) %>% 
  dplyr::ungroup()
table
#--------------------------------------------#

```



```{r table1-kable, appendix=TRUE}
# table 1 kable ------------------------#
knitr::kable(table, booktabs = TRUE,
             col.names = c("Species", "Sex", "Average body mass (g)"),
             align = 'llc',
             digits = 1,
             format.args = list(big.mark = ","),
             caption = "Palmer penguins' average body mass in grams by species and sex.",
             label = "table:kable")
#--------------------------------------------#

```

```{r table1-kableExtra, appendix=TRUE}
# table 1 kableExtra ------------------------#
knitr::kable(table %>% dplyr::select(-species), booktabs = TRUE,
             col.names = c("Species and sex", "Average body mass (g)"),
             align = 'lc',
             digits = 1,
             format.args = list(big.mark = ","),
             caption = "Palmer penguins' average body mass in grams by species and sex.",
             label = "table:kableExtra") %>% 
  kableExtra::pack_rows(index = table(table$species))
#--------------------------------------------#

```

The labels are important because you can refer to the tables like so:
compare Table \ref{table:kable} with Table \ref{table:kableExtra}.

Check out https://bookdown.org/yihui/rmarkdown-cookbook/tables.html for more tips!
Also the documentation for `kableExtra` here: https://haozhu233.github.io/kableExtra/awesome_table_in_pdf.pdf.

\FloatBarrier
\newpage



# Other interesting tips

You can print values of R variables in text.

```{r avg-body-mass, appendix=TRUE}
# calculate avg body mass ----------------#
avg_body_mass <- mean(penguins$body_mass_g, na.rm = TRUE)
#-----------------------------------------#

```

The average body mass in grams of all penguins is `r avg_body_mass`.
You can also use `round` and `scales::comma` to make it better-looking, 
so that it prints like so:
the average body mass in grams of all penguins is `r scales::comma(round(avg_body_mass,0))`.

You can find more tips chunk tips in here: https://yihui.org/knitr/options/.



\FloatBarrier
\newpage
# Appendix: Code

```{r all-code, ref.label=all_labels(appendix == TRUE), echo=TRUE, eval=FALSE}
```