---
title: "Lab 1 Demo"
author: "Chloe You"
date: "2023-01-13"
output:
  html_document:
    df_print: paged
  word_document: default
subtitle: General Workflow and Emphasis on Dataframes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

1. Create new project?
2. Check working directory. `getwd()` and `setwd()`
3. Housekeeping code: load libraries, load data

Short code for R script chunk

Windows : Ctrl + Alt + I

Mac : Command + Option + I


```{r, load-data, warning=FALSE, message=FALSE, include = FALSE}
# install.packages("gapminder")
library(gapminder)
library(dplyr)
## if local csv file
# dat <- read.csv("data/data.name.csv") # ?read.csv
dat <- gapminder # in our case we load it directly from package
```

Examine data!

## Demo 2
```{r, include=FALSE}
dplyr::glimpse(dat) 
str(dat)
#View(dat) ###Look at the entire dataset
head(dat)
tail(dat)
dim(dat)
nrow(dat)
ncol(dat)
sum(is.na(dat$country)) # missing data

dat
```
Inline code can be called as `r colnames(dat)`. The key is wrapping your R code in `r ` before putting them in your text. 

The mean life expectancy is `r round(mean(dat$lifeExp),2)`

We can also get the summary statistics and knit them into the report directly:
```{r}
knitr::kable(summary(dat),digits = 2, caption = "Summary Statistics")
```

## Demo 3 for Dataframes

Some useful ways to access cells in dataframes:
```{r}
dat[5,] # [row, column]

# dat$country
dat$country[2] # second row in the country column

```
Or we can (and we encourage you to) use the `dplyr` way to manipulate dataframes:

To extract rows which correspond to the year 2007 and select only the variables `country` and `pop` using the `dplyr` package:
```{r}
library(dplyr)
filter(dat, year == 2007) %>% 
  select(country, pop) # select columns

###Ctrl + Shift + m  %>% 
```

Finally we can try plotting the data too (with captions in the report too!)
```{r boxplot, fig.cap= "Life expectancy by continent"}
boxplot(lifeExp ~ continent , data = dat)
```


```{r scatterplot, fig.cap= "Life expectancy by log GDP per capita, coloured by continent"}
plot(lifeExp ~ log(gdpPercap), col=continent, dat=dat, xlab = "GDP per Capita",
     ylab = "Life Length")
```

And before you shut off RStudio, remember to save workspace so you can always come back to the same data, models next session!


